\documentclass[11pt,]{article}
\usepackage[margin=1in]{geometry}
\newcommand*{\authorfont}{\fontfamily{phv}\selectfont}
\usepackage[]{mathpazo}
\usepackage{abstract}
\renewcommand{\abstractname}{}    % clear the title
\renewcommand{\absnamepos}{empty} % originally center
\newcommand{\blankline}{\quad\pagebreak[2]}

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}} 
\usepackage{longtable,booktabs}

\usepackage{parskip}
\usepackage{titlesec}
\titlespacing\section{0pt}{12pt plus 4pt minus 2pt}{6pt plus 2pt minus 2pt}
\titlespacing\subsection{0pt}{12pt plus 4pt minus 2pt}{6pt plus 2pt minus 2pt}
\titleformat*{\subsubsection}{\normalsize\itshape}

\usepackage{titling}
\setlength{\droptitle}{-.25cm}

%\setlength{\parindent}{0pt}
%\setlength{\parskip}{6pt plus 2pt minus 1pt}
%\setlength{\emergencystretch}{3em}  % prevent overfull lines 

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}

\usepackage{fancyhdr}
\pagestyle{fancy}
\usepackage{lastpage}
\renewcommand{\headrulewidth}{0.3pt}
\renewcommand{\footrulewidth}{0.0pt} 
\lhead{}
\chead{}
\rhead{\footnotesize POLI210: Political Science Research Methods -- Fall
2021}
\lfoot{}
\cfoot{\small \thepage/\pageref*{LastPage}}
\rfoot{}

\fancypagestyle{firststyle}
{
\renewcommand{\headrulewidth}{0pt}%
   \fancyhf{}
   \fancyfoot[C]{\small \thepage/\pageref*{LastPage}}
}

%\def\labelitemi{--}
%\usepackage{enumitem}
%\setitemize[0]{leftmargin=25pt}
%\setenumerate[0]{leftmargin=25pt}




\makeatletter
\@ifpackageloaded{hyperref}{}{%
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
}
\@ifpackageloaded{color}{
    \PassOptionsToPackage{usenames,dvipsnames}{color}
}{%
    \usepackage[usenames,dvipsnames]{color}
}
\makeatother
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={ ()},
             pdfkeywords = {},  
            pdftitle={POLI210: Political Science Research Methods},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

% took this out in order to have numbered sections
% \setcounter{secnumdepth}{0} 





\usepackage{setspace}

\title{POLI210: Political Science Research Methods}
\author{Olivier Bergeron-Boutin}
\date{Fall 2021}


\begin{document}  

		\maketitle
		
	
		\thispagestyle{firststyle}

%	\thispagestyle{empty}


	\noindent \begin{tabular*}{\textwidth}{ @{\extracolsep{\fill}} lr @{\extracolsep{\fill}}}


E-mail: \texttt{\href{mailto:olivier.bergeron-boutin@mail.mcgill.ca}{\nolinkurl{olivier.bergeron-boutin@mail.mcgill.ca}}} & Class Hours: TR
8:35am-9:55am\\
Office: Zoom & 
Office Hours: T 10-11am; W 4-5pm; F 10-11am\\

	\vspace{1mm} \\
	\hline
	\end{tabular*}

\noindent \begin{tabular*}{\textwidth}{ @{\extracolsep{\fill}} lr @{\extracolsep{\fill}}}
TA: Marion Dessalles &
E-mail: \texttt{\href{mailto:marion.dessalles@mail.mcgill.ca}{\nolinkurl{marion.dessalles@mail.mcgill.ca}}}\\
Office hours: W 4-5 &
Office: Online \\
\hline
\end{tabular*}

\noindent \begin{tabular*}{\textwidth}{ @{\extracolsep{\fill}} lr @{\extracolsep{\fill}}}
	TA: Irene Zhang &
	E-mail: \texttt{\href{mailto:peiyi.zhang@mail.mcgill.ca}{\nolinkurl{peiyi.zhang@mail.mcgill.ca}}}\\
	Office hours: F 11:30-12:30 &
	Office: Online \\
	\hline
\end{tabular*}

\noindent \begin{tabular*}{\textwidth}{ @{\extracolsep{\fill}} lr @{\extracolsep{\fill}}}
	TA: Christopher Ross &
	E-mail: \texttt{\href{mailto:christopher.ross4@mail.mcgill.ca}{\nolinkurl{christopher.ross4@mail.mcgill.ca}}}\\
	Office hours: T 1-2 &
	Office: Leacock 520 \\
	\hline
\end{tabular*}

\noindent \begin{tabular*}{\textwidth}{ @{\extracolsep{\fill}} lr @{\extracolsep{\fill}}}
	TA: Danielle Bohonos &
	E-mail: \texttt{\href{mailto:danielle.bohonos2@mail.mcgill.ca}{\nolinkurl{danielle.bohonos2@mail.mcgill.ca}}}\\
	Office hours: F 1:30-2:30 &
	Office: Online \\
	\hline
\end{tabular*}
	
\vspace{2mm}
	


\hypertarget{what-is-this-course-for}{%
\section{What is this course for?}\label{what-is-this-course-for}}

This course is an introduction to research methods in political science.
It will cover a broad range of topics, including questions of philosophy
of science; causality and the difficult enterprise of causal inference;
how to ask empirical questions about the political world (and which ones
to ask!); formulating empirically testable hypotheses; measurement and
data collection using surveys and interviews; and qualitative and
quantitative data analysis techniques designed to test your hypotheses.

\hypertarget{pre-requisites-formal-and-informal}{%
\subsection{Pre-requisites -- formal and
informal}\label{pre-requisites-formal-and-informal}}

Formal: There are no pre-requisites for enrolling in this course.
However, if you have already taken POLI311, you are not eligible to take
this course. For those who matriculated in the fall of 2017 or later,
POLI210 serves as a prerequisite for POLI311.

Informal: You do \emph{not} need to have a math/stats background to
succeed in this course. We will provide you with the tools to succeed
regardless of your background.

\hypertarget{learning-outcomes}{%
\subsection{Learning outcomes}\label{learning-outcomes}}

\begin{itemize}
\tightlist
\item
  Become critical consumers of political science research and empirical
  claims in popular media.
\item
  Acquire knowledge of important quantitative and qualitative empirical
  methods in the study of politics.
\item
  Become acquainted with \emph{R} -- a free, open-source software that
  is highly marketable (this very syllabus and all course slides were
  created using \emph{R}!)
\end{itemize}

\hypertarget{textbook-and-other-resources}{%
\section{Textbook and other
resources}\label{textbook-and-other-resources}}

\hypertarget{textbook}{%
\subsection{Textbook}\label{textbook}}

The textbook for this course will be:

Berdahl, Loleen and Jason Roy. \emph{Explorations: Conducting Empirical
Research in Canadian Political Science} (4th edition). Oxford University
Press.

You can purchase the print version at the McGill bookstore or rent a
copy at \href{https://www.redshelf.com}{redshelf}. Note that we will be
using the newer 4th edition. If you use an older edition of the
textbook, it is incumbent upon you to make sure that you are not missing
important material.

\hypertarget{other-resources}{%
\subsection{Other resources}\label{other-resources}}

People learn in different ways and this is \emph{especially true} when
it comes to research methods. What works for me may not work for you!
Fortunately, there are a lot of resources you can use:

\begin{itemize}
\tightlist
\item
  StackOverflow is a Q\&A forum for programmers. If you type a question
  related to \emph{R} in Google, chances are that a StackOverflow thread
  will show up!
\item
  Thomas J. Leeper taught a similar course at LSE (the content of this
  very course is inspired by his). You can view the syllabus and slides
  on \href{https://thomasleeper.com/designcourse/}{his personal
  website}.
\item
  \emph{Quantitative Social Science: An Introduction} by Kosuke Imai is
  an excellent textbook that can help you learn \emph{R} and principles
  of statistics.
\item
  \href{https://r4ds.had.co.nz/index.html}{\emph{R for Data Science}} is
  a free online book that introduces you to the basics of
  \emph{R}/\texttt{tidyverse}.
\end{itemize}

\hypertarget{remote-delivery}{%
\section{Remote delivery}\label{remote-delivery}}

\hypertarget{lectures}{%
\subsection{Lectures}\label{lectures}}

The lectures will be held remotely through Zoom. They will be recorded
and posted on MyCourses.

\hypertarget{labs}{%
\subsection{Labs}\label{labs}}

The labs will be held in person. You are expected to be in Montreal for
the semester, but I understand that some may be unable/uncomfortable
attending labs in person. As such, the labs will be recorded and posted
on MyCourses. You are nonetheless encouraged to attend in person.

\hypertarget{quizzes-and-midterm}{%
\subsection{Quizzes and midterm}\label{quizzes-and-midterm}}

The quizzes and the midterm will be available on MyCourses. See the
``Evaluation policy'' section for more information.

\hypertarget{labs-r-and-rstudio}{%
\section{Labs: R and RStudio}\label{labs-r-and-rstudio}}

In labs, which are expected to take place in person, you will be
introduced to quantitative data analysis using the statistical software
\emph{R}, a free and open-source program. We will also be using RStudio,
which is a graphical user interface that makes learning R easier. You
can download \emph{R} \href{http://www.cran.r-project.org}{here}; note
that you need to choose the version that matches your operating system.
You can download RStudio
\href{https://www.rstudio.com/products/rstudio/download/}{here}. Try to
do this as soon as possible! You will need \emph{R} to complete most of
the assignments. Your presence in labs is essential to your success in
this course -- this is a skill that can only be learned by doing!

\hypertarget{optional-video-clips}{%
\section{Optional video clips}\label{optional-video-clips}}

In an effort to make the course materials more tangible, I will
occasionally produce short (and hopefully fun) videos that approach
things from a more informal angle. These are completely optional! This
course -- and other methods courses -- is an excellent gateway into
deeper, philosophical questions about the nature of knowledge. The aim
is to ease your entry into the world of empirical political science by
covering fun, real-world topics that would usually not be covered in a
course like this. Some of the topics we may cover include: recent
controversies in political science, the history of philosophy of
science, how norms vary across disciplines in the social sciences, how
the publication process works, what an efficient data analysis workflow
looks like\ldots Please feel free to suggest a different topic you would
like to hear about!

\hypertarget{mycourses-discussions}{%
\section{MyCourses discussions}\label{mycourses-discussions}}

I have set up a class discussion on \emph{MyCourses} (accessible under
the ``Discussions'' tab). You can ask general questions about the course
content, questions about the assignments, etc. \emph{NEVER} post
\emph{R} code! I have also set up a feedback discussion that you can use
if you want to provide any feedback on the course. You are welcome to
post anonymously if you prefer.

\hypertarget{evaluation-policy}{%
\section{Evaluation policy}\label{evaluation-policy}}

\begin{itemize}
\tightlist
\item
  Assignments: 45\%. There will be 5 assignments; assignment 1 is worth
  5 points; assignments 2 through 5 are worth 10 points each. You are
  allowed to talk generally about the problem sets with your classmates,
  but you are expected to complete them individually. \emph{You cannot
  share code with your classmates}.

  \begin{itemize}
  \tightlist
  \item
    Assignment 1, due on September 20th at 11:59pm, will simply ask to
    fill in a class survey on the platform \emph{Qualtrics}. The data
    you provide here will be of use to create some interactive examples
    in class!
  \item
    Assignment 2, due on October 7th at 11:59pm, will ask you to perform
    basic operations in \emph{R} using the class survey: compute
    measures of central tendency, recoding variables, describing
    distributions.
  \item
    Assignment 3, due on October 25th at 11:59pm, will ask you to work
    with experimental data from a famous political science article. It
    will address issues of causality.
  \item
    Assignment 4, due on November 1st at 11:59pm, will focus on the
    case-study and comparative approaches.
  \item
    Assignment 5, due on November 26th at 11:59pm, will ask you to work
    with data from the 2019 Canadian Election Study and will focus on
    comparisons between means, p-values, multivariate regression.
  \end{itemize}
\item
  Quizzes: 15\%. The first quiz will be posted to the ``Quiz'' section
  on MyCourses on Monday October 25th and you will need to complete it
  by Wednesday October 27th, 11:59pm. The second quiz will be posted on
  Tuesday November 30th and you will need to complete it by Thursday
  December 2nd, 11:59pm. Each is worth 7.5\%. Anything covered in
  lectures or in the readings is fair game; the format is multiple
  choice questions. Once you have opened the quiz, you will have 1.5
  hour to complete it.
\item
  Midterm: 20\%. The midterm will be posted on MyCourses on Tuesday
  November 9th and you will need to submit your answers by Friday
  November 12th, 11:59pm. Everything up to the material from week 10
  (inclusively) is fair game. You are allowed to use any course
  materials, but you are expected to complete the midterm on an
  individual basis. Once you have opened the midterm, you will have 3
  hours to complete it.
\item
  Team project: 20\%. Due on December 6th at 11:59pm. In teams of four,
  you will choose a political science article from a pre-approved list
  that will span the subfields of the discipline as well as the
  methodological approaches we will have covered in this course. You
  will first have to answer basic questions about the article: what is
  the research question? What is the hypothesis? Then, you will be asked
  to take a more critical approach and assess the methodological merits
  of the article. What are the main threats to inference and how well do
  the authors defend their research design? More details will be given
  in class in November.
\end{itemize}

\hypertarget{submissions}{%
\section{Submissions}\label{submissions}}

Everything will be handled remotely. These are the instructions for each
component:

\begin{itemize}
\tightlist
\item
  Quizzes: The quizzes will be available on MyCourses under the tab
  ``Quizzes.''
\item
  Midterm: The midterm will be posted on MyCourses and you will be able
  too upload the finished product under the ``Assignments'' tab.
\item
  Assignments: You will be able to upload your work under the
  ``Assignments'' tab. For the assignments involving R, you need to
  upload both your R script (.R file) as well as your written answers
  (in a PDF or Word document).
\end{itemize}

\hypertarget{course-and-university-policies}{%
\section{Course and University
policies}\label{course-and-university-policies}}

\hypertarget{lanuage-of-submission}{%
\subsection{Lanuage of submission}\label{lanuage-of-submission}}

In accord with McGill University's Charter of Student Rights, students
in this course have the right to submit in English or in French any
written work that is to be graded. This does not apply to courses in
which acquiring proficiency in a language is one of the objectives.''
(Approved by Senate on 21 January 2009)

\hypertarget{academic-integrity}{%
\subsection{Academic integrity}\label{academic-integrity}}

McGill University values academic integrity. Therefore, all students
must understand the meaning and consequences of cheating, plagiarism and
other academic offences under the Code of
\href{https://www.mcgill.ca/secretariat/files/secretariat/code_of_student_conduct_and_disciplinary_procedures.pdf}{Student
Conduct and Disciplinary Procedures}.

\hypertarget{copyright}{%
\subsection{Copyright}\label{copyright}}

© Instructor-generated course materials (e.g., handouts, notes,
summaries, exam questions) are protected by law and may not be copied or
distributed in any form or in any medium without explicit permission of
the instructor. Note that infringements of copyright can be subject to
follow up by the University under the Code of Student Conduct and
Disciplinary Procedures.

\hypertarget{late-submission}{%
\subsection{Late submission}\label{late-submission}}

Late assignments will be penalized 5 percentage points (out of 100) per
day. Late quizzes will not be accepted. Exceptions apply for illness,
family emergency, or other extraordinary circumstances. You will need to
provide documentation of your predicament and will be given more time to
complete your assignment/quiz.

\hypertarget{re-grading}{%
\subsection{Re-grading}\label{re-grading}}

Students who wish to contest a grade for an assignment or exam must do
so in writing (by email, sent to the instructor) providing the reasoning
behind their challenge to the grade received within two weeks of the day
on which the assignments are returned. The TA who graded the assignment
will re-grade your assignment, and may raise or lower the grade. If you
are still unsatisfied after the re-assessment, you can re-submit the
assignment to me (original copy with TA comments), along with your
justification. I will then re-evaluate the assignment, but also reserve
the right to raise or lower the grade. Please also see
\href{http://www.mcgill.ca/politicalscience/files/politicalscience/assessment_and_\%20re-read_policy_final.pdf}{the
department's guidelines}.

\hypertarget{class-schedule}{%
\section{Class schedule}\label{class-schedule}}

\textbf{Week 1}: Intro

\setlength{\leftskip}{1cm}

\begin{itemize}
\item
  Lecture 1.1 (Sept 2nd): Reviewing the syllabus and course components.
  Differentiating between two types of social scientific research
  (normative vs empirical). Three types of inference in empirical
  research (causal, descriptive, predictive).
\item
  Readings: None.
\end{itemize}

\setlength{\leftskip}{0cm}

\textbf{Week 2}: The foundations of scientific research

\setlength{\leftskip}{1cm}

\begin{itemize}
\item
  Lecture 2.1 (Sept 7th): We will cover some foundational questions:
  What makes science scientific? What are the principles of scientific
  research?
\item
  Lecture 2.2 (Sept 9th): The steps in the scientific process.
  Quantitative and qualitative approaches. How a scientist behaves.
\item
  Readings:

  \begin{itemize}
  \tightlist
  \item
    \emph{Explorations} chapters 1.
  \item
    \href{https://fivethirtyeight.com/features/the-supreme-court-is-allergic-to-math/}{The
    Supreme Court is Allergic to Math}, FiveThirtyEight.
  \item
    \href{https://statmodeling.stat.columbia.edu/2021/03/12/the-social-sciences-are-useless-so-why-do-we-study-them-heres-a-good-reason/}{The
    social sciences are useless. So why do we study them?}, by Andrew
    Gelman.
  \item
    Noel, Hans. 2010. ``Ten Things Political Scientists Know That You
    Don't.'' The Forum 8(3).
  \end{itemize}
\end{itemize}

\setlength{\leftskip}{0cm}

\textbf{Week 3}: Building theory and making hypotheses

\setlength{\leftskip}{1cm}

\begin{itemize}
\item
  Lecture 3.1 (Sept 14th): What is a scientific ``literature''? How do
  we come up with a research question, and what makes for a good one?
  How to build theory: inductive vs deductive approaches. Formulating
  hypotheses. Hypothesis testing and refining theory.
\item
  Lecture 3.2 (Sept 16th): Library workshop.
\item
  Readings:

  \begin{itemize}
  \tightlist
  \item
    \emph{Explorations} chapters 2 (full) and 4 (pp.55-58).
  \item
    Mutz, Diana C., and Jahnavi S. Rao. 2018. ``The Real Reason Liberals
    Drink Lattes.'' \emph{PS: Political Science \& Politics} 51(4):
    762--67.
  \end{itemize}
\end{itemize}

\setlength{\leftskip}{0cm}

\textbf{Week 4}: Causality

\setlength{\leftskip}{1cm}

\begin{itemize}
\item
  Lecture 4.1 (Sept 21st): Introduction to the potential outcomes
  framework and the fundamental problem of causal inference. Thinking
  counterfactually. Causal vs correlational relationships.
\item
  Lecture 4.2 (Sept 23rd): How random assignment solves the problem of
  self-selection. Experimental methods in political science. Internal vs
  external validity. The search for quasi-experiments.
\item
  Readings:

  \begin{itemize}
  \tightlist
  \item
    Druckman, James N., Donald P. Greene, James H. Kuklinski, and Arthur
    Lupia. 2011. ``Experiments.'' In \emph{Cambridge Handbook of
    Experimental Political Science}, Cambridge: Cambridge University
    Press, 15--26.
  \item
    \emph{Explorations} chapter 9.
  \item
    Butler, Daniel M., and David E. Broockman. 2011. ``Do Politicians
    Racially Discriminate Against Constituents? A Field Experiment on
    State Legislators.'' \emph{American Journal of Political Science}
    55(3): 463--77.
  \end{itemize}
\end{itemize}

\setlength{\leftskip}{0cm}

\textbf{Week 5}: Conceptualization and measurement

\setlength{\leftskip}{1cm}

\begin{itemize}
\item
  Lecture 5.1 (Sept 28th): Why do we even need concepts? How do define
  concepts? How do we differentiate one concept from another? The ladder
  of generality and ``definitional gerrymandering''
\item
  Lecture 5.2 (Sept 30th): 4 levels of measurement. Measurement
  obstacles: random and non-random measurement error; social
  desirability bias. Measurement validity.
\item
  Readings:

  \begin{itemize}
  \tightlist
  \item
    \emph{Explorations} chapter 4 (pp.58-66).
  \item
    Choose one:

    \begin{itemize}
    \tightlist
    \item
      Collier, David, and Steven Levitsky. 1997. ``Democracy with
      Adjectives: Conceptual Innovation in Comparative Research.''
      \emph{World Politics} 49(3): 430--51.
    \item
      Mansbridge, Jane. 2003. ``Rethinking Representation.''
      \emph{American Political Science Review} 97(4): 515--28.
    \item
      Lueders, Hans, and Ellen Lust. 2018. ``Multiple Measurements,
      Elusive Agreement, and Unstable Outcomes in the Study of Regime
      Change.'' \emph{The Journal of Politics} 80(2): 736--41.
    \end{itemize}
  \end{itemize}
\end{itemize}

\setlength{\leftskip}{0cm}

\textbf{Week 6}: Case studies and the comparative method

\setlength{\leftskip}{1cm}

\begin{itemize}
\item
  Lecture 6.1 (Oct 5th): Two types of case studies: descriptive vs
  theory-testing. Why focus on a single case? Scope conditions and
  generalization to other cases. What sort of evidence do we use?
\item
  Lecture 6.2 (Oct 7th): Goals of the comparative method. How to select
  cases: purposive sampling. Mill's method of similarity and method of
  difference. Can we establish causal relations?
\item
  Readings:

  \begin{itemize}
  \tightlist
  \item
    \emph{Explorations} chapter 10.
  \item
    Geddes, Barbara. 1990. ``How the Cases You Choose Affect the Answers
    You Get: Selection Bias in Comparative Politics.'' Political
    Analysis 2: 131--50.
  \end{itemize}
\end{itemize}

\setlength{\leftskip}{0cm}

\textbf{Week 7}: Interviews and focus groups

\setlength{\leftskip}{1cm}

\begin{itemize}
\item
  Lecture 7.1 (Oct 15th): Guest lecture by Professor Tania Islas
  Weinstein.
\item
  Readings:

  \begin{itemize}
  \tightlist
  \item
    \emph{Explorations} chapter 7.
  \item
    Fujii, Lee Ann. 2018. Interviewing in social science research: a
    relational approach (Routledge): ch.~1 \& 4 (pp.~1-11 \& 53-72).
  \end{itemize}
\end{itemize}

\setlength{\leftskip}{0cm}

\textbf{Week 8}: Ethnography and the interpretive method + survey
research I.

\setlength{\leftskip}{1cm}

\begin{itemize}
\item
  Lecture 8.1 (Oct 19th): Ethnography/interpretive method.
\item
  Lecture 8.2 (Oct 21st): A short history of political polling. How do
  we sample for polls?
\item
  Readings:

  \begin{itemize}
  \tightlist
  \item
    \emph{Explorations} chapter 5.
  \end{itemize}
\end{itemize}

\setlength{\leftskip}{0cm}

\textbf{Week 9}: Survey research II + descriptive statistics.

\setlength{\leftskip}{1cm}

\begin{itemize}
\item
  Lecture 9.1 (Oct 26th): The promises and pitfalls of survey research.
  What happened in 2016 and 2020 -- are polls broken?
\item
  Lecture 9.2 (Oct 28th): Measures of central tendency and measures of
  variance. Confidence intervals. The central importance of
  distributions.
\item
  Readings:

  \begin{itemize}
  \tightlist
  \item
    \emph{Explorations} chapter 8.
  \end{itemize}
\end{itemize}

\setlength{\leftskip}{0cm}

\textbf{Week 10}: The basics of statistical inference and hypothesis
testing.

\setlength{\leftskip}{1cm}

\begin{itemize}
\item
  Lecture 10.1 (Nov 2nd): Measures of association for different types of
  variables. Linear vs non-linear relationships.
\item
  Lecture 10.2 (Nov 4th): Difference-in-means and hypothesis testing.
  p-values and what they do/don't tell you.
\item
  Readings:

  \begin{itemize}
  \tightlist
  \item
    \emph{Explorations} chapter 12.
  \end{itemize}
\end{itemize}

\setlength{\leftskip}{0cm}

\textbf{Week 11}: Research ethics and the replication crisis

\setlength{\leftskip}{1cm}

\begin{itemize}
\item
  Lecture 11.1 (Nov 9th): Research ethics and the replication crisis.
  Advances in research transparency: public data, pre-registration,
  registered reports.
\item
  Lecture 11.2 (Nov 11th): No class (complete midterm).
\item
  Readings:

  \begin{itemize}
  \tightlist
  \item
    \emph{Explorations} chapter 3.
  \item
    Lee, Stephanie M. 2018.
    ``\href{https://www.buzzfeednews.com/article/stephaniemlee/brian-wansink-cornell-p-hacking\#.qs6A0K3DW}{Here's
    How Cornell Scientist Brian Wansink Turned Shoddy Data Into Viral
    Studies About How We Eat.}'' \emph{Buzzfeed News}.
  \item
    Singal, Jesse. 2015.
    ``\href{https://www.thecut.com/2015/05/how-a-grad-student-uncovered-a-huge-fraud.html}{The
    Case of the Amazing Gay-Marriage Data: How a Graduate Student
    Reluctantly Uncovered a Huge Scientific Fraud.}'' \emph{New York
    Magazine}.
  \end{itemize}
\end{itemize}

\setlength{\leftskip}{0cm}

\textbf{Week 12}: Linear regression

\setlength{\leftskip}{1cm}

\begin{itemize}
\item
  Lecture 12.1 (Nov 16th): The least squares algorithm and its
  mechanics.
\item
  Lecture 12.2 (Nov 18th): Interpreting regression output. Substantive
  vs statistical significance.
\item
  Readings:

  \begin{itemize}
  \tightlist
  \item
    \emph{Explorations} chapter 13.
  \end{itemize}
\end{itemize}

\setlength{\leftskip}{0cm}

\textbf{Week 13}: More on regression + mixed methods.

\setlength{\leftskip}{1cm}

\begin{itemize}
\item
  Lecture 13.1 (Nov 23rd): Basic regression diagnostics. What linear
  regression does and doesn't tell you about the political world. A
  short word on mixed methods.
\item
  Lecture 13.2 (Nov 25th): Preparation for the final project: how should
  we read empirical political science articles? How do we tell whether
  we should trust the findings?
\item
  Readings:

  \begin{itemize}
  \tightlist
  \item
    \emph{Explorations} chapter 14 (optional).
  \end{itemize}
\end{itemize}

\setlength{\leftskip}{0cm}

\textbf{Week 14}: Wrapping up

\setlength{\leftskip}{1cm}

\begin{itemize}
\item
  Lecture 14.1 (Nov 30th): Review session.
\item
  Lecture 14.2 (Dec 2nd): Empirical political science today -- what
  \emph{haven't} we learned in this class?
\item
  Readings:

  \begin{itemize}
  \tightlist
  \item
    None.
  \end{itemize}
\end{itemize}




\end{document}

\makeatletter
\def\@maketitle{%
  \newpage
%  \null
%  \vskip 2em%
%  \begin{center}%
  \let \footnote \thanks
    {\fontsize{18}{20}\selectfont\raggedright  \setlength{\parindent}{0pt} \@title \par}%
}
%\fi
\makeatother